Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-11-26 paper/manuscript.tex
revised abstract to <= 150 words

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt,lineno]{elife}
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Shifting mutational effects on the evolutionary landscape of HIV Envelope}

@@

\begin{abstract}
HIV's envelope protein (Env) evolves rapidly.
The immediate evolutionary space accessible to any viral variant is largely determined by the effects of single amino-acid mutations on viral fitness.
Due to epistasis, these effects can "shift" over evolutionary time as the virus traverses through sequence space (e.g., the effect of a mutation tolerated in one viral variant may shift such that it is not tolerated in another related variant).
In principle, even single amino-acid mutations can substantially alter a protein's mutational tolerance.
However, the prevalence of such shifts in long-term protein evolution in nature is largely unknown.
Here, we comprehensively quantified the ways in which the effects of mutations to Env have shifted or remained constant among two divergent Env homologs with 85\% sequence identity, corresponding to $>$100 amino-acid differences.
We did so by experimentally measuring the effects of all single amino-acid mutations to each Env homolog on the ability of that homolog to support viral replication in cell culture.
In total, we analyzed these effects at a total of 616 homologous sites.
We observed that a small fraction of mutations had largely shifted in their effects, where a given mutation was well tolerated in one homolog, but strongly disfavored in the other.
In contrast, we observed that the majority of mutations had shifted in their effects by only small-to-intermediate amounts.
Thus, our data suggest that mutational effects are substantially conserved between these divergent Env homologs despite $>$100 underlying amino-acid differences.
Overall, our data indicate that although epistasis can cause the effects of mutations to shift over evolutionary time, large shifts have been rare in the long-term evolutionary period separating the Env homologs we considered.
\end{abstract}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ELIFE ARTICLE TEMPLATE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PREAMBLE 
\documentclass[9pt]{elife}
% add lineno for line numbering
% Use the onehalfspacing option for 1.5 line spacing
% Use the doublespacing option for 2.0 line spacing
% Please note that these options may affect formatting.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Mapping mutational effects along the evolutionary landscape of HIV envelope}

@@

\begin{abstract}
The immediate evolutionary space accessible to a virus is largely determined by how single mutations affect fitness.
These mutational effects can shift as the virus evolves.
However, the prevalence of such shifts remains unclear.
Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) variants that differ at $>$100 residues.
Most mutations similarly affect both Envs, but the amino-acid preferences of a substantial minority of sites have clearly shifted.
These shifts usually involve sites that prefer a single amino acid in one Env but tolerate many amino acids in the other.
Surprisingly, shifts are only slightly enriched at sites that differ between the Envs -- and many occur at residues that do not even contact substitutions.
Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the evolutionary capacity of most sites is conserved between moderately diverged viral strains.
\end{abstract}
